Schema Evolution in Interactive Programming Systems

### 2024-08-24 paper.tex
Reviewer C nitpicks

% Inquiry: What problem or question does the paper address? How has this problem or question been addressed?
\noindent\textbf{Inquiry:} A key obstacle to live and collaborative programming systems is the problem of \emph{schema evolution}.
When a programmer changes the structure of a datatype, a programming system needs to co-evolve
existing live data and adapt code that implements the program behaviour to match the new structure.
This problem has been studied in isolation as schema evolution in databases, but also recurs in hot code reloading in live programming and dynamic software updating.
We connect these independent efforts in an interdisciplinary inquiry into the problem of schema change through all the layers of the development stack and across all the time-scales of programming.
%Such schema evolutions must occur immediately during live programming, but also neeed to be conveyed through long-term unplanned collaborations.

@@

The specific case where the type is refactored to a semantically equivalent type is related to
the Extract Entity challenge discussed below. Finally, a programming system tackling this
challenge may rely on structure editing so that the system has access to a high-level logical
description of the edits performed by the user.

@@

\subsection*{Remarks: Requirements and Representation}
It should be possible to automatically merge any two sequences of edits (changing both
the document structure and data) and the order in which the edits are applied should also not
matter. This may not always be possible and the system may ask the user for resolution or,
possibly, use a default behavior specified by the user.

@@

\section{Divergence Control in Spreadsheets}
Schema evolution becomes more challenging when data, code, and Schema cannot all be updated
together in an atomic way. In these situations, different versions must coexist and interoperate
with one another. For example, edits done to data of a new schema must be applied to
data of an older schema, or code written against older schema must run against data of a newer schema.

@@

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code. \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accomodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}. BiDEL appears capable of handling the basic requirements of our \emph{Extract/Absorb Entity} and \emph{Multiplicity Change} challenges, though \emph{Split/Merge Entity} is less clear. It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.

@@

\paragraph{Live programming.}
Live programming~\cite{tanimoto90, rein2018exploratory} seeks to speed up feedback in (primarily solo) programming.
% \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
\emph{Hot reloading/swapping}~\cite{barenz2020essence, hicks2005dynamic} installs code changes into a running system preserving some runtime state but not attempting to migrate that state through type changes.
\citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
\citet{RuntimeConstraint} study the live modeling problem in \S\ref{sec:livemodeling}. They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution. \citet{vanRozen19} take an alternative approach to the same problem using \emph{origin tracking} in textual editing.
%\citet{vanRozen23} subsequently proposed a metaprogramming language for live DSL programming.

>>>>

% Inquiry: What problem or question does the paper address? How has this problem or question been addressed?
\noindent\textbf{Inquiry:} A key obstacle to live and collaborative programming systems is the problem of \emph{schema evolution}.
When a programmer changes the structure of a datatype, a programming system needs to co-evolve
existing live data and adapt code that implements the program behaviour to match the new structure.
This problem has been studied in isolation as schema evolution in databases, but also recurs in hot code reloading in live programming and dynamic software updating.
We connect these independent efforts in an interdisciplinary inquiry into the problem of schema change through all the layers of the development stack and across all the time-scales of programming.
%Such schema evolutions must occur immediately during live programming, but also need to be conveyed through long-term unplanned collaborations.

@@

The specific case where the type is refactored to a semantically equivalent type is related to
the Extract Entity challenge discussed below. Finally, a programming system tackling this
challenge may rely on structure editing~\cite{Teitelbaum81} so that the system has access to a high-level logical
description of the edits performed by the user.

@@

\subsection*{Remarks: Requirements and Representation}
It should be possible to automatically merge any two sequences of edits (changing both
the document structure and data) and the order in which the edits are applied should also not
matter. One approach is to use CRDTs~\cite{Shapiro11} which guarantee order-independent convergence over a restricted set of edit operations. This may not always be possible and the system may ask the user for resolution or,
possibly, use a default behavior specified by the user.

@@

\section{Divergence Control}

Schema evolution becomes more challenging when data, code, and schema cannot all be updated
together in an atomic way. In these situations, different versions must coexist and interoperate
with one another. For example, edits done to data of a new schema must be applied to
data of an older schema, or code written against older schema must run against data of a newer schema.

@@

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to manage schema migration but is comparable to the capabilities of SQL DDL, often requiring the addition of custom Ruby or SQL code. Schema evolution is related to the classical view-update problem~\cite{Bancilhon81} which was approached from a language perspective by lenses~\cite{Foster2007}. Generalizing from lenses \citet{curino08} spawned a stream of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accomodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17}. BiDEL appears capable of handling the basic requirements of our \emph{Extract/Absorb Entity} and \emph{Multiplicity Change} challenges, though \emph{Split/Merge Entity} is less clear. It provides schema divergence by supporting multiple schema within one database, but would need some extension to handle data divergence.

@@

\paragraph{Live programming.}
Live programming~\cite{tanimoto90, rein2018exploratory, HazelnutLive19} seeks to speed up feedback in (primarily solo) programming.
% \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
\emph{Hot reloading/swapping}~\cite{barenz2020essence, hicks2005dynamic} installs code changes into a running system preserving some runtime state but not attempting to migrate that state through type changes.
\citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
\citet{RuntimeConstraint} study the live modeling problem in \S\ref{sec:livemodeling}. They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution. \citet{vanRozen19} take an alternative approach to the same problem using \emph{origin tracking} in textual editing.
%\citet{vanRozen23} subsequently proposed a metaprogramming language for live DSL programming.

### 2024-08-24 revisions.tex
Reviewer C nitpicks

In the introduction we clarify that this paper is not about a tool under development. Some of the problems are extracted from our different research projects but none of these projects span the rull range of problems.

@@

In \S\emph{Remarks: Managing Design Tradeoffs} we noted the importance of oferring solutions simple enough to be actually used.

@@

@@

Reviewer C
-----------------------------------------------------------------

- Live Programming for the Elm Architecture

I agree with the presented arguments but have difficulty understanding the remarks about requirements and implementation. The authors say that the migration needs to be done automatically but do not suggest a solution to the problem. Is it a type-based program synthesis? Generation of lenses? In what scenario does it need to be done automatically? A more straightforward way (but maybe not what the authors were thinking) would be to ask the programmer to write the transformation function. So, what is the goal of the idealised system being thought over?

A reference like https://hazel.org/ should be referenced here.

- Entity Evolution in Data-centric Systems

In this case, the authors miss a classic reference to the problem of updatable views, which deal with the same category of issues when solving a different problem. Migrations here feel a lot like the back translation problem in updatable views.

Also, this is the most common type of evolution problem in the software industry, which is treated by disciplining the software process, versioning, and creating adapters. A contrast with these solutions, pointing again to the ghost system, would be very welcome. This approach is presented in section 6, but it makes sense to refer here as well.

https://www.irisa.fr/lande/lande/icse-proceedings/seams/p86.pdf

- Evolution of Computational Documents

This is a problem that existing approaches solve relatively well, and there is work using, for instance, operation-based CRDTs to solve the problem of collaborative editing and ensure eventual consistency. I think those solutions should be referred to, and their results should be compared to the goals presented in this section.

- Divergence control in Spreadsheets

The problem in this use case is broader than spreadsheets, and the section title may be misleading. Versioning schemes, like semantic versioning or language-based versioning, could solve the problem of divergence control in spreadsheets. This class of solutions could help present this case better.

https://dl.acm.org/doi/10.1145/3479394.3479416
Multiplicity Change in Data Formats (Migrations)

Using lenses to solve this problem would be a very good idea to highlight here, as in some old work of Benjamin Pierce and a system called Unison. This problem was central in transformations and sync operations in XML documents. I have no specific reference to point out, but Unison seems a good place to start. The synthesis of lenses can be done.

- Live Modelling Languages

This case study is, in my opinion, the more solid one in terms of presentation. It almost seems that this is the idealised system that the authors are thinking about. This presentation also gives away the identity of the authors (Programming is not double-blind; I am just referring to this since the authors omitted their names in the paper).

The comment about the presented example being out of reach for all existing language workbenches could be expanded to explain why. Are there any techniques extracted from the remaining case studies that would be useful here?

- Overall

the discussion and conclusions are very brief both at the end and in each case study and do not provide a clear path to solving the challenges presented in the paper. Even a speculative section pointing to combining features from different systems would be welcome to communicate the authors' vision of the future of interactive programming systems.

>>>>

In the introduction we clarify that this paper is not about a tool under development. Some of the problems are extracted from our different research projects but none of these projects span the full range of problems.

@@

In \S\emph{Remarks: Managing Design Tradeoffs} we noted the importance of offering solutions simple enough to be actually used.

@@

in \S\emph{Remarks: Requirements and Implementation} we clarify that the goal is to adapt to schema changes automatically to avoid interrupting live programming.

We added to Related Work the history of how the view-update problem led to lenses which led to DELs.

We cited Hazel in Related Work.

We now mention the possibility of using CRDTs for evolving computational documents.

Divergence Control for Spreadsheets has been corrected to just Divergence Control.

As mentioned above, we have added an explanation in the introduction of why we maintain neutrality by avoiding speculation on our preferred solutions.

@@
